A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

### 2020-05-10 Manuscript/Manuscript-6-DCA.Rmd
added paragraph mentioned by Dean

>>>>

This study is not the first to compare $\lambda$ and $\kappa$ for their ability as statistics to measure phylogenetic signal.  Our results for $\lambda$ and $\kappa$ values are consistent with those found in the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], but that study investigated type I error rates and statistical power, finding that $\lambda$ performed better in both regards, irrespective of species number in trees.  Although not the central focus of their study, the same tendency for variable $\lambda$ and consistent $\kappa$ at intermediate phylogenetic signal strengths was observed [see Fig. 2, @Munkemuller_et_al2012].  Recent work by Molina-Venegas and Rodríguez [-@MolinaVenegas2017] found that $\kappa$ but not $lambda$ tended to inflate the estimate of phylogenetic signal, leading to moderate type I and type II biases, if polytomic chronograms were used.  Their work more thoroughly addressed previous observations of inflated $\kappa$ for incompletely resolved phylogenetic trees [@Davies2012; @Munkemuller_et_al2012].  An interesting question is whether an inflated $\kappa$ value leads to an inflated $Z_\kappa$ or does a tendency of a particular tree to inflate estimates of $\kappa$ also inflate the values in random permutations of a test, in which case $Z_\kappa$ is robust to polytomies?  We repeated the analyses in Figure 4, adjusting trees to have 50% collapsed nodes, per the technique of Molina-Venegas and Rodríguez [-@MolinaVenegas2017], and found results were consistent (see Supporting Information).  This confirms that any tendency of incompletely resolved trees to inflate $\kappa$ as a descriptive statistic does not inflate $Z_\kappa$ as an effect size, and because comparison of effect sizes in a test is a comparison of locations of observed values in their sampling distributions, if the sampling distributions also shift because of this tendency, the $Z_{12}$ test statistic in equation 4 appears to be robust in spite of unresolved trees.   \hfill\break

***Note from Mike*** I did not use Boettiger's blog as ammo because he also speculates that $\kappa$ is problematic and I did not want to point a gun at our bow.  I think we need to add a section to the SI, which I will do, but I did a quick simulation using Molina-Venegas and Rodríguez's method for making a tree polytomy-rich and compared $\kappa$ and $Z$ from physignal analyses for $\lambda = 0$ and $\lambda = 1$ to the plots in Fig. 4 and there is no difference.  So I wrote the paragraph before having the full results but will obtain those results in a daya or two.  Take-home point, any distortion of $kappa$ because of trees does not translate to distortion of $Z_{\kappa}$!   \hfill\break

### 2020-05-10 Manuscript/Manuscript-6-DCA.tex
added paragraph mentioned by Dean

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Munkemuller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The (\(\lambda\)) parameter transforms the lengths of the
internal branches of the phylogeny to improve a fit of data to the
phylogeny via maximum likelihood (Pagel 1999; Freckleton et al. 2002).
Pagel's \(\lambda\) ranges from \(0\to1\), with larger values signifying
a greater dependence of observed trait variation on the phylogeny.
Pagel's \(\lambda\) also has the appeal that it may be included in
phylogenetic generalized least-squares regression (PGLS) to account for
the degree of phylogenetic signal in comparative analyses (see
Freckleton et al. 2002). \hfill\break

@@

\hypertarget{the-precision-of-lambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The Precision of \(\lambda\) is
Variable}}{The Precision of \textbackslash lambda is Variable}}\label{the-precision-of-lambda-is-variable}}

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
published in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

@@

Based on the effect size \(Z_\kappa\), we then proposed a two-sample
test, which provides means of determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another, via a hypothesis test. Prior studies have summarized patterns
of variation in phylogenetic signal across datasets using summary test
values, such as \(\kappa\) (e.g., Blomberg et al. 2003). However,
\(\kappa\) does not scale linearly with input levels of phylogenetic
signal, and its variance increases (i.e., precision decreases) with
increasing strength of phylogenetic signal (Munkemuller et al. 2012;
Diniz-Filho et al. 2012: see also Supporting Information). Thus,
\(\kappa\) should not be considered an effect size that measures the
strength of phylogenetic signal on a common scale. By contrast,
standardizing \(\kappa\) (\(Z_\kappa\), via equation 2) alleviates these
concerns, and facilitates formal statistical comparisons of the strength
of signal across datasets. Thus when viewed from this perspective, the
approach developed here aligns well with other statistical approaches
such as meta-analysis (sensu Hedges and Olkin 1985; Glass 1976; Arnqvist
and Wooster 1995), where summary statistics across datasets are
converted to standardized effect sizes for subsequent ``higher order''
statistical summaries or comparisons. As such our approach enables
evolutionary biologists to quantitatively examine the relative strength
of phylogenetic signal across a wide range of phenotypic traits, and
thus opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.
\hfill\break

@@

@@

@@

\leavevmode\hypertarget{ref-MolinaVenegas2017}{}%
Molina-Venegas, R., and M. A. Rodriguez. 2017. Revisiting phylogenetic
signal; strong or negligible impacts of polytomies and branch length
information? BMC evolutionary biology 17:53.

\leavevmode\hypertarget{ref-Munkemuller_et_al2012}{}%
Munkemuller, T., S. Lavergne, B. Bzeznik, S. Dray, T. Jombart, K.
Schiffers, and W. Thuiller. 2012. How to measure and test phylogenetic
signal. Methods in Ecology and Evolution 3:743--756.

@@

\leavevmode\hypertarget{ref-Orme2013}{}%
Orme, D., R. P. Freckleton, G. H. Thomas, T. Petzoldt, S. A. Fritz, and
N. Isaac. 2013. CAPER: Comparative analyses of phylogenetics and
evolution in r. Methods in Ecology and Evolution 3:145--151.

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets, including measures of serial independence
(\(\mathbf{C}\): Abouheif 1999), autocorrelation estimates (\(I\):
Gittleman and Kot 1990), statistical ratios of trait variation relative
to what is expected given the phylogeny (\(\kappa\): Blomberg et al.
2003; Adams 2014a), and scaling parameters used in maximum likelihood
fitting of the data to the phylogeny (\(\lambda\): Pagel 1999), among
others (e.g., Klingenberg and Gidaszewski 2010). The statistical
properties of these methods -- namely type I error rates and power --
have also been investigated to determine when phylogenetic signal can be
detected and under what conditions (e.g., Münkemüller et al. 2012;
Pavoine and Ricotta 2012; Diniz-Filho et al. 2012; Adams 2014a;
Molina-Venegas and Rodríguez 2017; see also Revell et al. 2008; Revell
2010). One of the most widely used methods for characterizing
phylogenetic signal in macroevolutionary studies is Pagel's \(\lambda\)
(Pagel 1999). The (\(\lambda\)) parameter transforms the lengths of the
internal branches of the phylogeny to improve a fit of data to the
phylogeny via maximum likelihood (Pagel 1999; Freckleton et al. 2002).
Pagel's \(\lambda\) ranges from \(0\to1\), with larger values signifying
a greater dependence of observed trait variation on the phylogeny.
Pagel's \(\lambda\) also has the appeal that it may be included in
phylogenetic generalized least-squares regression (PGLS) to account for
the degree of phylogenetic signal in comparative analyses (see
Freckleton et al. 2002). \hfill\break

@@

\hypertarget{the-precision-of-lambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The Precision of \(\lambda\) is
Variable}}{The Precision of \textbackslash{}lambda is Variable}}\label{the-precision-of-lambda-is-variable}}

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al. (2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
published in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

@@

Based on the effect size \(Z_\kappa\), we then proposed a two-sample
test, which provides means of determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another, via a hypothesis test. Prior studies have summarized patterns
of variation in phylogenetic signal across datasets using summary test
values, such as \(\kappa\) (e.g., Blomberg et al. 2003). However,
\(\kappa\) does not scale linearly with input levels of phylogenetic
signal, and its variance increases (i.e., precision decreases) with
increasing strength of phylogenetic signal (Münkemüller et al. 2012;
Diniz-Filho et al. 2012: see also Supporting Information). Thus,
\(\kappa\) should not be considered an effect size that measures the
strength of phylogenetic signal on a common scale. By contrast,
standardizing \(\kappa\) (\(Z_\kappa\), via equation 2) alleviates these
concerns, and facilitates formal statistical comparisons of the strength
of signal across datasets. Thus when viewed from this perspective, the
approach developed here aligns well with other statistical approaches
such as meta-analysis (sensu Hedges and Olkin 1985; Glass 1976; Arnqvist
and Wooster 1995), where summary statistics across datasets are
converted to standardized effect sizes for subsequent ``higher order''
statistical summaries or comparisons. As such our approach enables
evolutionary biologists to quantitatively examine the relative strength
of phylogenetic signal across a wide range of phenotypic traits, and
thus opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.
\hfill\break

@@

This study is not the first to compare \(\lambda\) and \(\kappa\) for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \(\kappa\) values are consistent with those found in
the simulations performed by Münkemüller et al. (2012), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \(\kappa\) at
intermediate phylogenetic signal strengths was observed (see Fig. 2,
Münkemüller et al. 2012). Recent work by Molina-Venegas and Rodríguez
(2017) found that \(\kappa\) but not \(lambda\) tended to inflate the
estimate of phylogenetic signal, leading to moderate type I and type II
biases, if polytomic chronograms were used. Their work more thoroughly
addressed previous observations of inflated \(\kappa\) for incompletely
resolved phylogenetic trees (Davies et al. 2012; Münkemüller et al.
2012). An interesting question is whether an inflated \(\kappa\) value
leads to an inflated \(Z_\kappa\) or does a tendency of a particular
tree to inflate estimates of \(\kappa\) also inflate the values in
random permutations of a test, in which case \(Z_\kappa\) is robust to
polytomies? We repeated the analyses in Figure 4, adjusting trees to
have 50\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (2017), and found results were consistent (see Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \(\kappa\) as a descriptive statistic does not inflate
\(Z_\kappa\) as an effect size, and because comparison of effect sizes
in a test is a comparison of locations of observed values in their
sampling distributions, if the sampling distributions also shift because
of this tendency, the \(Z_{12}\) test statistic in equation 4 appears to
be robust in spite of unresolved trees. \hfill\break

\textbf{\emph{Note from Mike}} I did not use Boettiger's blog as ammo
because he also speculates that \(\kappa\) is problematic and I did not
want to point a gun at our bow. I think we need to add a section to the
SI, which I will do, but I did a quick simulation using Molina-Venegas
and Rodríguez's method for making a tree polytomy-rich and compared
\(\kappa\) and \(Z\) from physignal analyses for \(\lambda = 0\) and
\(\lambda = 1\) to the plots in Fig. 4 and there is no difference. So I
wrote the paragraph before having the full results but will obtain those
results in a daya or two. Take-home point, any distortion of \(kappa\)
because of trees does not translate to distortion of \(Z_{\kappa}\)!
\hfill\break

@@

\leavevmode\hypertarget{ref-Davies2012}{}%
Davies, T. J., N. J. Kraft, N. Salamin, and E. M. Wolkovich. 2012.
Incompletely resolved phylogenetic trees inflate estimates of
phylogenetic conservatism. Ecology 93:242--247. Wiley Online Library.

@@

\leavevmode\hypertarget{ref-MolinaVenegas2017}{}%
Molina-Venegas, R., and M. A. Rodríguez. 2017. Revisiting phylogenetic
signal; strong or negligible impacts of polytomies and branch length
information? BMC evolutionary biology 17:53.

\leavevmode\hypertarget{ref-Munkemuller_et_al2012}{}%
Münkemüller, T., S. Lavergne, B. Bzeznik, S. Dray, T. Jombart, K.
Schiffers, and W. Thuiller. 2012. How to measure and test phylogenetic
signal. Methods in Ecology and Evolution 3:743--756.

@@

\leavevmode\hypertarget{ref-Orme2013}{}%
Orme, D., R. Freckleton, G. Thomas, T. Petzoldt, S. Fritz, and N. Isaac.
2013. CAPER: Comparative analyses of phylogenetics and evolution in r.
Methods in Ecology and Evolution 3:145--151.
